Think deep for reef resilience

### 2025-08-25 paper/paper_draft.qmd
Add caveat around not including the influence of light/depth on growth
Something in this commit seems to have broken the VSCode visual editor line wrapping setting

---
title: Think deep for reef resilience
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
    attributes:
        corresponding: true
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great Barrier
    Reef (GBR) experiencing severe declines in coral cover from thermal stress events.
    Understanding drivers behind reef resilience to climate impacts is critical for conservation
    planning and intervention strategies. A scenario-based population modelling approach was
    adopted with larval connectivity dynamics, thermal adaptation, and environmental factors
    to assess coral reef resilience across GBR locations under future climate projections.
    Using five Global Climate Models (GCMs) spanning five Equilibrium Climate Sensitivities,
    we analysed reef cover time series and the ability of reefs to maintain positive carbonate
    production budgets under future conditions. Contrary to expectations, larval connectivity
    patterns did not correlate with increases in maintenance of positive carbonate budgets.
    Instead, reef depth emerged as the primary predictor of cover-time series resilience and
    maintenance of positive carbonate budgets, with deeper reefs benefiting from thermal
    refugia that reduce bleaching impacts. Our findings suggest that depth provides critical
    protection against thermal stress, contrasting current restoration practices that primarily
    target easily accessible shallow sites. These results have important implications for
    reef management, indicating that depth refugia should be prioritized for further
    consideration in conservation planning alongside connectivity considerations to maximize
    coral survival under continuing climate change.
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
  custom:
  - kind: float
    reference-prefix: STable
    key: stbl
editor:
  markdown:
    wrap: 92
---

@@

| Global Climate Model source | ECS value (°C)     | Interpreted relative severity | Results found in |
|-----------------------|-----------------------|------------------------|-----------------------|
| ACCESS-CM2                  | 4.66 [@grose2023]  | High                          | [@grier2025]     |
| EC-Earth3-Veg               | 4.30 [@wyser2020]  | High                          | [@grier2025]     |
| GFDL-CM4                    | 4.1 [@sentman2025] | High                          | [@grier2025]     |
| ACCESS-ESM1-5               | 3.87 [@ziehn2020]  | Intermediate                  | Main text        |
| NorESM2-MM                  | 2.5 [@seland2020]  | Low                           | [@grier2025]     |

@@

There are multiple caveats in how coral ecosystem dynamics are modelled in this
investigation. We model coral colonies as circular diameter values that are tracked over
timesteps through growth. Due to the availability of quality future projected data we do not
consider other important impacts such as waves, cyclones, currents and Crown of Thorns
Starfish (CoTS) on coral colony survival and bleaching [@massel1993; @deaker2022].
Uncertainty in connectivity modelling outputs, satellite-derived bathymetry data and future
thermal stress projections are likely to influence the results obtained in this
investigation. While the resolution of connectivity data used in this investigation improves
upon data used in comparable studies [@bozec2025], the time-mean implementation used may not
effectively capture long term patterns [@vogt-vincent2023]. Using alternative connectivity
implementation methods that better capture the ecological-timescale trends in reef
connections rather than the high daily variability, would improve confidence in the
connectivity findings of this investigation [@vogt-vincent2023]. The DHW data used in this
investigation have not been bias corrected for the Great Barrier Reef region and are part of
a wider Australian dataset [@dixon2022] developed using statistical downscaling.
Additionally, Lachs et al. [@lachs2021] suggest the utility of DHW levels for prediction of
coral bleaching can be improved by reducing the threshold for HotSpot Anomaly values from
MMM +1 to MMM and reducing the accumulation window from 12 weeks to 8 weeks. We investigate
a range of future climate responses, thermal projections and ecological parameterisations of
ADRIA - CoralBlox to account for many possible futures, and our results suggest the factors
that influence reef resilience are consistent across these scenarios.

@@

The results presented suggest that when modelled as part of a large and complex ecosystem,
reef depth is a key factor contributing to ongoing reef resilience. We provide evidence in
support of depth as a form of thermal refugia for shallow-water reefs on the GBR. While
protecting ‘key node’ reefs in larval connectivity networks is an important consideration
for reef restoration efforts [@anthony2020; @hock2017], current connectivity modelling
methods, in combination with our modelling framework, suggest that connectivity may not be
as important for reef resilience under projected conditions as environmental factors such as
depth and thermal stress levels. We suggest further analyses that investigate the benefits
of incorporating depth-based resilience into reef intervention decision making, as well as
further research and development in connectivity modelling methods for whole-ecosystem
scales.

@@

>>>>

---
title: Think deep for reef resilience
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
    attributes:
        corresponding: true
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great Barrier
    Reef (GBR) experiencing severe declines in coral cover from thermal stress events.
    Understanding drivers behind reef resilience to climate impacts is critical for conservation
    planning and intervention strategies. A scenario-based population modelling approach was
    adopted with larval connectivity dynamics, thermal adaptation, and environmental factors
    to assess coral reef resilience across GBR locations under future climate projections.
    Using five Global Climate Models (GCMs) spanning five Equilibrium Climate Sensitivities,
    we analysed reef cover time series and the ability of reefs to maintain positive carbonate
    production budgets under future conditions. Contrary to expectations, larval connectivity
    patterns did not correlate with increases in maintenance of positive carbonate budgets.
    Instead, reef depth emerged as the primary predictor of cover-time series resilience and
    maintenance of positive carbonate budgets, with deeper reefs benefiting from thermal
    refugia that reduce bleaching impacts. Interaction between reef depth and incoming connectivity was found to influence reef temporal behaviour. Our findings suggest that depth provides critical
    protection against thermal stress, contrasting current restoration practices that primarily
    target easily accessible shallow sites. These results have important implications for
    reef management, indicating that depth refugia should be prioritized for further
    consideration in conservation planning alongside connectivity considerations to maximize
    coral survival under continuing climate change.
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
  custom:
  - kind: float
    reference-prefix: STable
    key: stbl
editor:
  markdown:
    wrap: 92
---

@@

| Global Climate Model source | ECS value (°C) | Interpreted relative severity | Results found in |
|------------------|------------------|-------------------|------------------|
| ACCESS-CM2 | 4.66 [@grose2023] | High | [@grier2025] |
| EC-Earth3-Veg | 4.30 [@wyser2020] | High | [@grier2025] |
| GFDL-CM4 | 4.1 [@sentman2025] | High | [@grier2025] |
| ACCESS-ESM1-5 | 3.87 [@ziehn2020] | Intermediate | Main text |
| NorESM2-MM | 2.5 [@seland2020] | Low | [@grier2025] |

@@

There are multiple caveats in how coral ecosystem dynamics are modelled in this investigation. We model coral colonies as circular diameter values that are tracked over timesteps through growth. Additionally, we do not model the direct influence of light intensity on coral growth or the negative influence of depth on coral growth rates due to reduced light levels. \[**BG: Include citations for influence of depth/light on growth rates**\]. If included in this investigation, it is possible that reef recovery ability would be reduced at greater depths even though corals at these depths experience refugia from thermal stress. Due to the availability of quality future projected data we do not consider other important impacts such as waves, cyclones, currents and Crown of Thorns Starfish (CoTS) on coral colony survival and bleaching [@massel1993; @deaker2022].

Uncertainty in connectivity modelling outputs, satellite-derived bathymetry data and future thermal stress projections are likely to influence the results obtained in this investigation. While the resolution of connectivity data used in this investigation improves upon data used in comparable studies [@bozec2025], the time-mean implementation used may not effectively capture long term patterns [@vogt-vincent2023]. Using alternative connectivity implementation methods that better capture the ecological-timescale trends in reef connections rather than the high daily variability, would improve confidence in the connectivity findings of this investigation [@vogt-vincent2023]. The DHW data used in this investigation have not been bias corrected for the Great Barrier Reef region and are part of a wider Australian dataset [@dixon2022] developed using statistical downscaling. Additionally, Lachs et al. [@lachs2021] suggest the utility of DHW levels for prediction of coral bleaching can be improved by reducing the threshold for HotSpot Anomaly values from MMM +1 to MMM and reducing the accumulation window from 12 weeks to 8 weeks. We investigate a range of future climate responses, thermal projections and ecological parameterisations of ADRIA - CoralBlox to account for many possible futures, and our results suggest the factors that influence reef resilience are consistent across these scenarios.

@@

The results presented suggest that when modelled as part of a large and complex ecosystem, reef depth is a key factor contributing to ongoing reef resilience. We provide evidence in support of depth as a form of thermal refugia for shallow-water reefs on the GBR. While protecting ‘key node’ reefs in larval connectivity networks is an important consideration for reef restoration efforts [@anthony2020; @hock2017], current connectivity modelling methods, in combination with our modelling framework, suggest that connectivity may not be as solely-important for reef resilience under projected conditions as environmental factors such as depth and thermal stress levels. We suggest further analyses that investigate the benefits of incorporating depth-based resilience into reef intervention decision making, as well as further research and development in connectivity modelling methods for whole-ecosystem scales.

@@

# Acknowledgements

# Conflict of Interest Statement

# Data Availability Statement
